---
title: Introduction to R
source: Rmd
teaching: 30
exercises: 10
---

::::::::::::::::::::::::::::::::::::::: objectives

- Assign values to objects in R.
- Learn how to name objects.
- Use comments to inform script.
- Solve simple arithmetic operations in R.
- Call functions and use arguments to change their default options.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- What is an object?
- What is a function and how can we pass arguments to a function?
- How can values be initially assigned to variables of different data types?

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, include=FALSE}
```

## Creating objects in R

:::::::::::::::::::::::::::::::::::::::::  callout

## Objects vs. variables

What are known as `objects` in `R` are known as `variables` in many other
programming languages. Depending on the context, `object` and `variable` can
have drastically different meanings. However, in this lesson, the two words
are used synonymously. For more information see:
[https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Objects](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Objects)


::::::::::::::::::::::::::::::::::::::::::::::::::


We will start by typing a simple mathematical operation in the console:

```{r, purl=FALSE}
3 + 5
7 * 2   # multiply 7 by 2
sqrt(36) # take the square root of 36
```


However, to do useful and interesting things, we need to assign **values** to
`objects`. 

To create an object, we need to provide:
- a name (e.g. 'first_number')
- a value (e.g. '1')
- the assignment operator ('\<-')

```{r, purl=FALSE}
time_minutes <- 5        # assign the number 5 to the object time_minutes
```

:::::::::::::::::::::::::::::::::::::::::  callout

## The assignment operator `<-`

`<-` is the assignment operator. It assigns values on the right to objects on
the left. Here we are creating a symbol called `time_minutes` and assigning it
the numeric value 5. Some R users would say "`time_minutes` *gets* 5."
`time_minutes` is now a *numeric vector* with one element.

The ‘=’ works too, but is most commonly used in passing arguments to functions 
(more on functions later). There are shortcuts for the R assignment operator:
- Windows shortcut: <KBD>Alt</KBD>\+<KBD>\-</KBD>
- Mac shortcut: <KBD>Option</KBD>\+<KBD>\-</KBD>

:::::::::::::::::::::::::::::::::::::::::  


When you assign something to a symbol, nothing happens in the console, but in
the Environment pane in the upper right, you will notice a new object,
`time_minutes`.


:::::::::::::::::::::::::::::::::::::::::  callout

## Tips on naming objects

Here are some tips for naming objects in R:

- **Do not use names of functions that already exist in R:** There are some
  names that cannot be used because they are the names of fundamental functions in
  R (e.g., `if`, `else`, `for`, see
  [here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Reserved.html)
  for a complete list. In general, even if it's allowed, it's best to not use
  other function names (e.g., `c`, `T`, `mean`, `data`, `df`, `weights`). If in
  doubt, check the help to see if the name is already in use.
- **R is case sensitive**: `age` is different from `Age` and `y` is different
  from `Y`.
- **No blank spaces or symbols other than underscores**: Instead, use capitalization 
  (e.g. `myData`) or underscores (e.g. `my_data`). It's also best to avoid dots (`.`) 
  within an object name as in `my.dataset`. There are many functions in R with dots in 
  their names for historical reasons, but dots have a special meaning in R (for methods)
  and other programming languages.
- **Do not begin with numbers or symbols**: `2x` is not valid, but `x2` is.
- **Be descriptive, but keep the names short**: Be descriptive with your variable names. 
  If you're loading in a lot of data, choosing `myData` or `x` as a name may not be as 
  helpful as, say, `ebookUsage`.

:::::::::::::::::::::::::::::::::::::::::  


## Evaluating Expressions

If you now type `time_minutes` into the console, and press Enter on your keyboard, R will
*evaluate* the expression. In this case, R will *print* the elements that are
assigned to `time_minutes` (the number 5). We can do this easily since y only has one
element, but if you do this with a large dataset loaded into R, it will overload
your console because it will print the entire thing. The `[1]` indicates that
the number 5 is the first element of this vector.

When assigning a value to an object, R does not print anything to the console. You
can force R to print the value by using parentheses or by typing
the object name:

```{r, purl=FALSE}
time_minutes <- 5    # doesn't print anything
(time_minutes <- 5)  # putting parenthesis around the call prints the value of y
time_minutes         # so does typing the name of the object
print(time_minutes)  # so does using the print() function.
```

Now that R has `time_minutes` in memory, we can do arithmetic with it. For
instance, we may want to convert it into seconds (60 seconds in 1 minute):

```{r, purl=FALSE}
60 * time_minutes
```

We can also change an object's value by assigning it a new one:

```{r, purl=FALSE}
time_minutes <- 10
60 * time_minutes
```

This overwrites the previous value without prompting you, so be careful! Also,
assigning a value to one object does not change the values of other objects  For
example, let's store the time in seconds in a new object, `time_seconds`:

```{r, purl=FALSE}
time_seconds <- 60 * time_minutes
```

Then change `time_minutes` to 30:

```{r, purl=FALSE}
time_minutes <- 30
```

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

What do you think is the current content of the object `time_seconds`? 600 or
1800?

:::::::::::::::  solution

## Solution

The value of `time_seconds` is still 600 because you have not
re-run the line `time_seconds <- 60 * time_minutes` since
changing the value of `time_minutes`.



:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::


## Removing objects from the environment

To remove an object from your R environment, use the `rm()` function. Remove
multiple objects with `rm(list = c("add", "objects", "here))`, adding the
objects in `c()` using quotation marks. 

To remove all objects, use `rm(list = ls())` or click the broom icon in the 
Environment Pane, next to "Import Dataset."

```{r, purl=FALSE, eval=FALSE}
x <- 5
y <- 10
z <- 15
rm(x)  # remove x
rm(list =c("y", "z"))  # remove y and z
rm(list = ls())  # remove all objects
```

:::::::::::::::::::::::::::::::::::::::  callout

## Tip: Use whitespace for readability

The white spaces surrounding the assignment operator `<-` are unnecessary. 
However, including them does make your code easier to read. There are several 
style guides you can follow, and choosing one is up to you, but consistency is key!

A style guide we recommend is the [Tidyverse style guide](https://style.tidyverse.org/).

:::::::::::::::::::::::::::::::::::::::  


:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

Create two variables `my_length` and `my_width` and assign them any numeric values you want.
Create a third variable `my_area` and give it a value based on the the multiplication of `my_length` and `my_width`.
Show that changing the values of either `my_length` and `my_width` does not affect the value of `my_area`.

:::::::::::::::  solution

## Solution

```{r}
my_length <- 2.5
my_width <- 3.2
my_area <- my_length * my_width
area
# change the values of my_length and my_width
my_length <- 7.0
my_width <- 6.5
# the value of my_area isn't changed
my_area
```

:::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::



## Comments

All programming languages allow the programmer to include comments in their
code. To do this in R we use the `#` character. Anything to the right of the `#`
sign and up to the end of the line is treated as a comment and will not be
evaluated by R. You can start lines with comments or include them after any code
on the line.

Comments are essential to helping you remember what your code does, and
explaining it to others. Commenting code, along with documenting how data is
collected and explaining what each variable represents, is essential to
reproducible research. See the Software Carpentry lesson on [R for Reproducible
Scientific Analysis](https://swcarpentry.github.io/r-novice-gapminder/).

```{r}
time_minutes <- 5    # time in minutes
time_seconds <- 60 * time_minutes	# convert to seconds
time_seconds				# print time in seconds
```

RStudio makes it easy to comment or uncomment a paragraph: after selecting the
lines you  want to comment, press at the same time on your keyboard
<kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>C</kbd>. If you only want to comment
out one line, you can put the cursor at any location of that line (i.e. no need
to select the whole line), then press <kbd>Ctrl</kbd> + <kbd>Shift</kbd> +
<kbd>C</kbd>.


## Functions and their arguments

R is a "functional programming language," meaning it contains a number of
*functions* you use to do something with your data. Functions are "canned
scripts" that automate more complicated sets of commands. Many functions are
predefined, or can be made available by importing R *packages* as we saw in the
"Before We Start" lesson.

*Call* a function on a variable by entering the function into the console,
followed by parentheses and the variables. A function usually gets one or more
inputs called *arguments*. For example, if you want to take the sum of 3 and 4,
you can type in `sum(3, 4)`. In this case, the arguments must be a number, and
the return value (the output) is the sum of those numbers. An example of a
function call is:

```{r, eval=FALSE, purl=FALSE}
sum(3, 4)
```

The function `is.function()` will check if an argument is a function in R. If it
is a function, it will print `TRUE` to the console.

Functions can be nested within each other. For example, `sqrt()` takes the
square root of the number provided in the function call. Therefore you can run
`sum(sqrt(9), 4)` to take the sum of the square root of 9 and add it to 4.

Typing a question mark before a function will pull the help page up in the
Navigation Pane in the lower right. Type `?sum` to view the help page for the
`sum` function. You can also call `help(sum)`. This will provide the description
of the function, how it is to be used, and the arguments.

In the case of `sum()`, the ellipses `. . .` represent an unlimited number of
numeric elements.

```{r, eval=FALSE, purl=FALSE}
is.function(sum)        # check to see if sum() is a function
sum(3, 4, 5, 6, 7)      # sum takes an unlimited number (. . .) of numeric elements
```

### Arguments

Some functions take arguments which may either be specified by the user, or, if
left out, take on a *default* value. However, if you want something specific,
you can specify a value of your choice which will be used instead of the
default. This is called *passing* an argument to the function.

For example, `sum()` takes the argument option `na.rm`. If you check the help
page for sum (call `?sum`), you can see that `na.rm` requires a logical
(`TRUE/FALSE`) value specifying whether `NA` values (missing data) should be
removed when the argument is evaluated.

By default, `na.rm` is set to `FALSE`, so evaluating a sum with missing
values will return `NA`:

```{r, eval=TRUE, purl=FALSE}
sum(3, 4, NA)                # 
```

Even though we do not see the argument here, it is operating in the background,
as the `NA` value remains. 3 + 4 + `NA` is `NA`.

But setting the argument `na.rm` to `TRUE` will remove the `NA`:

```{r, eval=TRUE, purl=FALSE}
sum(3, 4, NA, na.rm = TRUE)
```

It is very important to understand the different arguments that functions take,
the values that can be added to those functions, and the default arguments.
Arguments can be anything, not only `TRUE` or `FALSE`, but also other objects.
Exactly what each argument means differs per function, and must be looked up in
the documentation.

It's good practice to put the non-optional arguments first in your function
call, and to specify the names of all optional arguments.  If you don't, someone
reading your code might have to look up the definition of a function with
unfamiliar arguments to understand what you're doing.

:::::::::::::::::::::::::::::::::::::::  discussion

## Discussion

Type in `?round` at the console and then look at the output in the Help pane.
What other functions exist that are similar to `round`?
How do you use the `digits` parameter in the round function?

::::::::::::::::::::::::::::::::::::::::::::::::::



:::::::::::::::::::::::::::::::::::::::: keypoints

- Use the assignment operator \<- to assign values to objects. 
- Remove existing objects using the `rm()` function.
- Add comments in R scripts using the `#` operator.
- Define and use R functions and arguments. Functions automate more complicated sets of commands. Many functions are predefined, or can be made available by importing R packages.

::::::::::::::::::::::::::::::::::::::::::::::::::


