---
title: Working with Vectors
source: Rmd
teaching: 35
exercises: 15
---

::::::::::::::::::::::::::::::::::::::: objectives

- Recognise the different data types in R.
- Create, inspect and manipulate the contents of a vector.
- Subset and extract values from vectors.
- Analyze vectors with missing data.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- What are the different data types in R?
- What is a vector? How can a vector be created?
- How can subsets be extracted from vectors?
- How does R treat missing values?
- How can we deal with missing values in R?

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, include=FALSE}
```


## Vectors and data types

```{r, echo=FALSE, purl=TRUE}
### Vectors and data types
```

A vector is the most common and basic data type in R, and is pretty much the
workhorse of R. A **vector** is a sequence of elements of the same *type*. Vectors
can only contain "homogenous" data--in other words, all data must be of the same
type. The type of a vector determines what kind of analysis you can do on it.
For example, you can perform mathematical operations on **numeric** objects, but
not on `character` objects.

We can assign a series of values to a vector using the `c()` function. `c()`
stands for combine. If you read the help files for `c()` by calling `help(c)`,
you can see that it takes an unlimited `. . .` number of arguments.

For example we can create a vector of checkouts for a collection of books and
assign it to a new object `checkouts`:

```{r, purl=FALSE}
checkouts <- c(25, 15, 18)
checkouts
```

A vector can also contain **characters**. For example, we can have
a vector of the book titles (`title`) and authors (`author`):

```{r, purl=FALSE}
title <- c("Macbeth","Dracula","1984")
```

:::::::::::::::::::::::::::::::::::::::: callout

## Tip: Using quotes

The quotes around "Macbeth", etc. are essential here. Without the quotes R will
assume there are objects called `Macbeth` and `Dracula` in the environment. As
these objects don't yet exist in R's memory, there will be an error message.

:::::::::::::::::::::::::::::::::::::::: 


### Inspecting a vector

There are many functions that allow you to inspect the content of a vector.

For instance, `length()` tells you how many elements are in a particular vector:

```{r, purl=FALSE}
length(checkouts)  # print the number of values in the checkouts vector
```

An important feature of a vector is that all of the elements are the same type
of data. The function `class()` indicates the class (the type of element) of an
object:

```{r, purl=FALSE}
class(checkouts)
class(title)
```

You can call `str()` on an R object to compactly display information about it,
including the data type, number of elements, and a printout of the first few
elements.

```{r, purl=FALSE}
str(checkouts)
str(title)
```

You can use the `c()` function to add other elements to your vector:

```{r, purl=FALSE}
author <- "Stoker"
author <- c(author, "Orwell") # add to the end of the vector
author <- c("Shakespeare", author)
author
```

:::::::::::::::::::::::::::::::::::::::: reminder

To know what a function does, type `?function_name` into the console. 
For example, you can type `?str` to read the description for the `str()` function.

:::::::::::::::::::::::::::::::::::::::: 


In the first line, we create a character vector `author` with a single value
`"Stoker"`. In the second line, we add the value `"Orwell"` to it, and save the
result back into `author`. Then we add the value `"Shakespeare"` to the
beginning, again saving the result back into `author`.

We can do this over and over again to grow a vector, or assemble a dataset.
As we program, this may be useful to add results that we are collecting or
calculating.

Vectors are one of the many **data structures** that R uses. Other important
ones are lists (`list`), matrices (`matrix`), data frames (`data.frame`),
factors (`factor`) and arrays (`array`).

:::::::::::::::::::::::::::::::::::::::: callout

An **atomic vector** is the simplest R **data type** and is a linear vector of a
single type. Above, we saw 2 of the 6 main **atomic vector** types  that R uses:
`"character"` and `"numeric"` (or `"double"`). These are the basic building
blocks that all R objects are built from. The other 4 **atomic vector** types
are:

- `"logical"` for `TRUE` and `FALSE` (the boolean data type)
- `"integer"` for integer numbers (e.g., `2L`, the `L` indicates to R that it's an integer)
- `"complex"` to represent complex numbers with real and imaginary parts (e.g.,
  `1 + 4i`) and that's all we're going to say about them
- `"raw"` for bitstreams that we won't discuss further

You can check the type of your vector using the `typeof()` function and
inputting your vector as the argument.

:::::::::::::::::::::::::::::::::::::::: reminder


:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

We've seen that atomic vectors can be of type character, numeric (or double),
integer, and logical. But what happens if we try to mix these types in a
single vector?

:::::::::::::::  solution

## Solution

R implicitly converts them to all be the same type.


:::::::::::::::::::::::::

What will happen in each of these examples? (hint: use `typeof()`
to check the data type of your objects):

```{r, results="hide"}
num_char <- c(1, 2, 3, "a")
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c("a", "b", "c", TRUE)
tricky <- c(1, 2, 3, "4")
```

Why do you think it happens?

:::::::::::::::  solution

## Solution

Vectors can be of only one data type. R tries to
convert (coerce) the content of this vector to find a "common
denominator" that doesn't lose any information.


:::::::::::::::::::::::::

How many values in `combined_logical` are `"TRUE"` (as a character) in the
following example:

```{r, results="hide"}
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c("a", "b", "c", TRUE)
combined_logical <- c(num_logical, char_logical)
```

 

:::::::::::::::  solution

## Solution

Only one. There is no memory of past data types, and the coercion
happens the
first time the vector is evaluated. Therefore, the `TRUE` in
`num_logical`
gets converted into a `1` before it gets converted into `"1"` in
`combined_logical`.


:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::


### Coercion

You've probably noticed that objects of different types get converted into a
single, shared type within a vector. In R, we call converting objects from one
class into another class **coercion**. These conversions happen according to a
hierarchy, whereby some types get preferentially coerced into other types: 
<kbd>logical<kbd> \< <kbd>integer<kbd> \< <kbd>numeric<kbd> \< <kbd>complex<kbd> \< 
<kbd>character<kbd> \< <kbd>list<kbd>.

You can also *coerce* a vector to be a specific data type with `as.character()`,
`as.logical()`, `as.numeric()`, etc. For example, to coerce a number to a
character:

```{r, purl=FALSE}
x <- as.character(200)
```

We can test this in a few ways: if we print `x` to the console, we see quotation
marks around it, letting us know it is a character:

```{r, purl=FALSE}
x
```

We can also call `class()`

```{r, purl=FALSE}
class(x)
```

And if we try to add a number to `x`, we will get an error message `non-numeric argument to binary operator`\--in other words, x is `non-numeric` and cannot be
added to a number.

```{r, purl=FALSE, eval=FALSE}
x + 5
```


## Subsetting vectors

If we want to subset (or extract) one or several values from a vector, we must
provide one or several indices in square brackets. For this example, we will use
the `state` data, which is built into R and includes data related to the 50
states of the U.S.A. Type `?state` to see the included datasets. `state.name` is
a built in vector in R of all U.S. states:

```{r, results="show", purl=FALSE}
state.name
state.name[1]
```

You can use the `:` colon to create a vector of consecutive numbers.

```{r, results="show", purl=FALSE}
state.name[1:5] 
```

If the numbers are not consecutive, you must use the `c()` function:

```{r, results="show", purl=FALSE}
state.name[c(1, 10, 20)]
```

We can also repeat the indices to create an object with more elements than the
original one:

```{r, results="show", purl=FALSE}
state.name[c(1, 2, 3, 2, 1, 3)]
```

R indices start at 1. Programming languages like Fortran, MATLAB, Julia, and R
start counting at 1, because that's what human beings typically do. Languages in
the C family (including C++, Java, Perl, and Python) count from 0 because that's
simpler for computers to do.


### Conditional subsetting

Another common way of subsetting is by using a logical vector. `TRUE` will
select the element with the same index, while `FALSE` will not:

```{r, results="show", purl=FALSE}
five_states <- state.name[1:5]
five_states[c(TRUE, FALSE, TRUE, FALSE, TRUE)]
```

Typically, these logical vectors are not typed by hand, but are the output of
other functions or logical tests. `state.area` is a vector of state areas in
square miles. We can use the `<` operator to return a logical vector with TRUE
for the indices that meet the condition:

```{r, results="show", purl=FALSE}
state.area < 10000
state.area[state.area < 10000]
```

The first expression gives us a logical vector of length 50, where `TRUE`
represents those states with areas less than 10,000 square miles. The second
expression subsets `state.name` to include only those names where the value is
`TRUE`.

You can also specify character values. `state.region` gives the region that each
state belongs to:

```{r, results="show", purl=FALSE}
state.region == "Northeast"
state.name[state.region == "Northeast"]
```

Again, a `TRUE/FALSE` index of all 50 states where the region is the Northeast,
followed by a subset of `state.name` to return only those `TRUE` values.

Sometimes you need to do multiple logical tests (think Boolean logic). You can
combine multiple tests using `|` (at least one of the conditions is true, OR) or
`&` (both conditions are true, AND). Use `help(Logic)` to read the help file.

```{r logical, comment=NA, results="asis", echo=F, eval=FALSE}
multiple <- data.frame("operator" = c("&", "|", "!", "any", "all")
                       , "function" = c("boolean AND", "boolean OR", "boolean NOT", "ANY true", "ALL true")
                       , stringsAsFactors = F)

kable(multiple) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

```{r, results="show", purl=FALSE}
state.name[state.area < 10000 | state.region == "Northeast"]
state.name[state.area < 10000 & state.region == "Northeast"]
```

The first result includes both states with fewer than 10,000 sq. mi. and all
states in the Northeast. New York, Pennsylvania, Delaware and Maine have areas
with greater than 10,000 square miles, but are in the Northeastern U.S. Hawaii
is not in the Northeast, but it has fewer than 10,000 square miles. The second
result includes only states that are in the Northeast and have fewer than 10,000
sq. mi.

R contains a number of operators you can use to compare values. Use
`help(Comparison)` to read the R help file. Note that **two equal signs** (`==`)
are used for evaluating equality (because one equals sign (`=`) is used for
assigning variables).

```{r logical1, comment=NA, results="asis", echo=F, eval=FALSE}
operators <- tibble("operator" = c("<", ">", "==", "<=", ">=", "!=", "%in%", "is.na", "!is.na"),
                    "function" = c("Less Than", "Greater Than", "Equal To", "Less Than or Equal To", "Greater Than or Equal To", "Not Equal To", "Has A Match In", "Is NA", "Is Not NA"))

kable(operators) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

A common task is to search for certain strings in a vector.  One could use the
"or" operator `|` to test for equality to multiple values, but this can quickly
become tedious. The function `%in%` allows you to test if any of the elements of
a search vector are found:

```{r, results="show", purl=FALSE}
west_coast <- c("California", "Oregon", "Washington")
state.name[state.name == "California" | state.name == "Oregon" | state.name == "Washington"]
state.name %in% west_coast
state.name[state.name %in% west_coast]
```


## Missing data

As R was designed to analyze datasets, it includes the concept of missing data
(which is uncommon in other programming languages). Missing data are represented
in vectors as `NA`. R functions have special actions when they encounter `NA.`

When doing operations on numbers, most functions will return `NA` if the data
you are working with include missing values. This feature makes it harder to
overlook the cases where you are dealing with missing data. As we saw above, you
can add the argument `na.rm=TRUE` to calculate the result while ignoring the
missing values.

```{r, purl=FALSE}
rooms <- c(2, 1, 1, NA, 4)
mean(rooms)
max(rooms)
mean(rooms, na.rm = TRUE)
max(rooms, na.rm = TRUE)
```

If your data include missing values, you may want to become familiar with the
functions `is.na()`, `na.omit()`, and `complete.cases()`. See below for
examples.

```{r, purl=FALSE}
## Use any() to check if any values are missing
any(is.na(rooms))

## Use table() to tell you how many are missing vs. not missing
table(is.na(rooms))

## Identify those elements that are not missing values.
complete.cases(rooms)

## Identify those elements that are missing values.
is.na(rooms)

## Extract those elements that are not missing values.
rooms[complete.cases(rooms)]
```

You can also use `!is.na(rooms)`, which is exactly the same as
`complete.cases(rooms)`. The exclamation mark indicates logical negation.

```{r, purl=FALSE}
!c(TRUE, FALSE)
```

How you deal with missing data in your analysis is a decision you will have to
make--do you remove it entirely? Do you replace it with zeros? That will depend
on your own methodological questions.

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

1. Using this vector of rooms, create a new vector with the NAs removed.
  
  ```r
  rooms <- c(1, 2, 1, 1, NA, 3, 1, 3, 2, 1, 1, 8, 3, 1, NA, 1)
  ```

2. Use the function `median()` to calculate the median of the `rooms` vector.

3. Use R to figure out how many households in the room variable have more than 2 rooms.

:::::::::::::::  solution

## Solution

```{r, answer=TRUE}
rooms <- c(1, 2, 1, 1, NA, 3, 1, 3, 2, 1, 1, 8, 3, 1, NA, 1)
rooms_no_na <- rooms[!is.na(rooms)]
# or
rooms_no_na <- na.omit(rooms)
# 2.
median(rooms, na.rm = TRUE)
# 3.
rooms_above_2 <- rooms_no_na[rooms_no_na > 2]
length(rooms_above_2)
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

Now that we have learned how to write scripts, and the basics of R's data
structures, we are ready to start working with the library catalog dataset and learn about data frames.



:::::::::::::::::::::::::::::::::::::::: keypoints

- Define the following terms as they relate to R: object, vector, assign, call, function.
- A vector is a sequence of elements of the same type. All data in a vector must be of the same type--character, numeric (or double), integer, and logical. 
- Create or add new objects to a vector using the c() function.
- Use \[ \] to subset values from vectors.
- Deal with missing data in vectors using the is.na(), na.omit(), and complete.cases() functions.

::::::::::::::::::::::::::::::::::::::::::::::::::


